---
title: "My course notes"
author: "Madelaine Norström"
date: "2024-11-18"
output:

  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-   We selected 20 isolates

    -   8 are presumptive ***Salmonella***
    -   6 are presumptive *Klebsiella*
    -   6 are presumptive *Escherichia coli*

    ## Installation of packages

    ```{r}

   # install.packages("here")
   # install.packages("DiagrammeR")
      


    ```

# ctrl + alt + I 


You can also embed plots, for example:
ctrl + alt +I ( make a cell - short cut on the keyboard)

```{r}
angry_or_not<-"not"
```
gives  the content 

Madelaine, are you angry_or_not? `r angry_or_not`

# working with your data

```{r}
here::here()

```

```{r}
path_data<-here::here("data","2024-09-25_INIKA_SAMPLING_ANALYSIS_HUMAN.csv")
```

```{r}
human_data<-
  readr::read_csv2(path_data)

```
#Day 3


## Packages into memory


```{r}
library(here)
library (tidyverse)
```


```{r}
?library()
```

## Exploration of a dataframe 
```{r}

colnames(human_data)
```
- overview of data and data types
```{r}
dput(
  colnames(human_data)
  )
```
- explanation: c() is to make vectors

## Getting help with packages and what does a function do

F1

```{r}
test<-c("one","2", 3)
test

typeof(test)
class(test)


```
All the elements of a vector are of the same type ( tesxt, double0REAL, integer, data,,,) 


```{r}
test2<-c(1,2,3)
test2

typeof(test2)
class(test2)


```
```{r}
test2*test2
```
```{r}
test2*c(1,2)

```

- explanation: c() is to make vectors

```{r}
test <- c("one", "2", 3)
test
typeof(test)
class(test)
```
- figure out if the character is numeric or text or something else
- what type of objects do we have?

# sign for an assignment: <-

```{r}
test <- c("one", "2", 3)
test
test[3]
typeof(test[3])

class(test)
```
All the elements of a vector are of the same `type`(text, double = real, integer, date ....)

```{r}
test2 <- c(1,2,3)
typeof(test2)
class(test2)
test2*test2
```

```{r}
test2 * c(1,2)
```

- deactivate the output in the markdown document; in the document settings near the knit button: only select `Chunk output in console`

- overview of the data and the data types in your data

```{r}
glimpse(human_data)
str(human_data)
```

- need to reduce the number of columns

- clean the data for spaces

```{r}
original_names <- colnames(human_data)
original_names
```

- replace ALL spaces with underscores

```{r}
library(tidyverse)
new_names <- str_replace_all(original_names, " ", "_")
new_names
new_names[88]
```
- regex pattern (regular expression)
- what criteria to use to choose what to exchange in your data? need to know your data and check your file while you do the replacements to see if it looks like you expect

```{r}
new_names2 <- str_replace_all(
  new_names, "[?/,;.*()-]", "_") %>% 
  str_remove_all("(^_*)|(_*$)")
new_names2
new_names2[88]
```

```{r}
colnames(human_data) <- new_names2
```

- learn how to select columns and transfer the dataset we need to reduce the size of the table

```{r}
human_data %>% 
  select(INIKA_OH_TZ_ID, Age__yrs, Gender, ) %>% 
  head()
```


```{r}
human_data_selection <- 
  human_data %>% 
  select(INIKA_OH_TZ_ID, Age__yrs, Gender, Enter_a_date, Region, District, 
         Specify_if_other_district, Sample, Season, Origin_of_sample, Which_class_grade_are_you, 
         Who_is_your_caretaker, 
         If_others__mention, 
         What_is_your_occupation_and_or_of_your_caretaker, 
         Have_you_ever_heard_about_AMR, If_yes__how_did_you_get_this_information, 
         Have_you_or_your_children_used_any_antibiotics_at_any_time, 
         If_yes__where_did_you_get_these_drugs_from,
         If_it_was_drug_sellers_or_pharmacy__did_you_have_a_prescription_from_the_doctor_prescriber,
        GPS_coordinates) 

human_data_selection %>% View()
human_data_selection %>% head()
```

<--should have done changes to the file before running the last command to remove _ at the end of the header text-->

new_columns <- 
  colnames(human_data) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all( "[?/,;.*()-]", "_") %>%
  str_remove_all("(^_*)|(_*$)")


- looking at the structure of your data

```{r}
glimpse(human_data_selection)
str(human_data_selection)
```

- remove empty columns or somethings that do not make sense

```{r}
human_data_selection <- 
  human_data_selection %>% 
  select(-Specify_if_other_district) %>% 
  mutate_at(vars(Age__yrs), as.integer) %>% 
  mutate_if(is.character, factor) %>% 
  mutate_at(vars(INIKA_OH_TZ_ID), as.character)
```

- mutate: want to change the data type for the INIKA_OH_TZ_ID from numerical to text/chr/integer
- categorical variable vs character (bivariable)
- when happy; reassign the code from human_data_selection %>% to <- and remove str() before ``` at the end of the code. We will pipe in the beginning when you work on the code to figure out what to change or remove in your data. 

- want a summary of the contents of the data:

```{r}
summary(
  human_data_selection
  )
```

- filter out values/data

```{r}
human_data_selection %>% 
  filter(INIKA_OH_TZ_ID == "238")
```

- delete the filtered value: 

```{r}
human_data_selection %>% 
  filter(INIKA_OH_TZ_ID != "238") %>% 
  filter(is.na(Gender))
  #summary() 
  #%>% 
  #View()
```









